---
title: "PTSD"
author: "Simon Schwab"
date: "14 Sep 2018"
output: html_notebook
---

## Libraries
```{r message=FALSE}
 # install.packages("ggplot2")
 # install.packages("cowplot")
 # install.packages("foreign")
 # install.packages("testit")
 # 
 # # sudo apt-get install libcurl4-openssl-dev
 # install.packages("car")
 # install.packages("reshape2")
 # install.packages("caret")

library(ggplot2)
library(cowplot)
library(foreign)
library(testit)
library(car)
library(reshape2)
library(caret)
library(scales)
library(knitr)
library(biostatUZH)
```

## Variables and paths
```{r}
PATH_HOME = path.expand('~')
PATH = file.path(PATH_HOME, 'Dropbox', 'Data', 'PTSD')
PATH_DATA = file.path(PATH, 'data')
PATH_FIGURES = file.path(PATH, 'figures')
```

## Load data
```{r warning=FALSE}
mydata = read.spss(file.path(PATH_DATA, 'PTSD_Skalen_1 06 2017_GESAMT_LS.sav'), 
                   to.data.frame = T)

N = nrow(mydata)

# Some ids are dublicates but in fact different subjects, thus we recreate the IDs
mydata$id_new = as.factor(1:N)
```

### Translate variables
```{r} 
mydata$occupation =  as.character(mydata$arbeitsort)
mydata$occupation[mydata$occupation=="Sanität, Ambulanz"] = "Ambulance service"
mydata$occupation[mydata$occupation=="Feuerwehr"] = "Fire service"
mydata$occupation[mydata$occupation=="Polizei"] = "Police"
mydata$occupation[mydata$occupation=="Spitalnotfall"] = "Emergency staff"
mydata$occupation[mydata$occupation=="Psychiatrie"] = "Psychiatry"
mydata$occupation = factor(mydata$occupation, ordered = F,
                               levels = c("Police", "Fire service", "Ambulance service",
                                          "Emergency staff", "Psychiatry"))
Ng = summary(mydata$occupation)
```

### Check variables
```{r eval=FALSE, include=FALSE}
summary(mydata$wohnsituation)
mydata$id[which(mydata$wohnsituation=="7")] # 4 cases have factor "7"

summary(mydata$schwierigesEreignis)
mydata$id[which(mydata$schwierigesEreignis=="4")] 
```

### Missings
```{r}
sum(is.na(mydata$SWE_Skala))/N
mydata$occupation.r[is.na(mydata$SWE_Skala)]
```

### Create variables

#### Support
```{r}
mydata$support = (as.numeric(mydata$hilfeBelastEinsatz_SQ001) +
                    as.numeric(mydata$hilfeBelastEinsatz_SQ012) +
                    as.numeric(mydata$hilfeBelastEinsatz_SQ013)) / 3
```

#### Belastung in Situationen

##### Ordinal
```{r}
# nie erlebt keine wenig etwas stark sehr stark (Ausmass der Belastung)
# NA         0     1     2     3     4
items = names(mydata)[grep('ereignisart_SQ0[0-9][0-9]$', names(mydata))]

for (i in 1:length(items)) {
  mydata[[paste(items[i], ".ord", sep = "")]] = rep(NA, N)
  mydata[[paste(items[i], ".ord", sep = "")]][mydata[[items[i]]]=='nie erlebt'] = NA
  mydata[[paste(items[i], ".ord", sep = "")]][mydata[[items[i]]]=='keine'] = 0
  mydata[[paste(items[i], ".ord", sep = "")]][mydata[[items[i]]]=='wenig'] = 1
  mydata[[paste(items[i], ".ord", sep = "")]][mydata[[items[i]]]=='etwas'] = 2
  mydata[[paste(items[i], ".ord", sep = "")]][mydata[[items[i]]]=='stark'] = 3
  mydata[[paste(items[i], ".ord", sep = "")]][mydata[[items[i]]]=='sehr stark'] = 4
}

mydata$belastung_menschen = rowMeans(cbind(
  mydata$ereignisart_SQ001.ord, mydata$ereignisart_SQ002.ord, mydata$ereignisart_SQ003.ord,
  mydata$ereignisart_SQ004.ord, mydata$ereignisart_SQ005.ord, mydata$ereignisart_SQ006.ord,
  mydata$ereignisart_SQ008.ord, mydata$ereignisart_SQ009.ord), na.rm = TRUE)


mydata$belastung_beruf = rowMeans(cbind(
  mydata$ereignisart_SQ007.ord, mydata$ereignisart_SQ010.ord, mydata$ereignisart_SQ011.ord,
  mydata$ereignisart_SQ012.ord, mydata$ereignisart_SQ013.ord), na.rm = TRUE)

# 3 missings
mydata$belastung_menschen[is.na(mydata$belastung_menschen)] = median(mydata$belastung_menschen, na.rm = T)
```

##### Categorial with only three levels
```{r}
# nie erlebt keine wenig etwas stark sehr stark (Ausmass der Belastung)
# NA         0     1     2     3     4
items = names(mydata)[grep('ereignisart_SQ0[0-9][0-9]$', names(mydata))]

for (i in 1:length(items)) {
  
  mydata[[paste(items[i], ".new", sep = "")]] = rep(NA, N)
  x = mydata[[items[i]]]
  mydata[[paste(items[i], ".new", sep = "")]][x == "nie erlebt"] = "never"
  mydata[[paste(items[i], ".new", sep = "")]][x == "keine" | x == "wenig" | x == "etwas"] = "little"
  mydata[[paste(items[i], ".new", sep = "")]][x == "stark" | x == "sehr stark"] = "strong"
  
  mydata[[paste(items[i], ".new", sep = "")]] = 
    factor(mydata[[paste(items[i], ".new", sep = "")]], levels = c("strong", "little", "never"))
}
```

##### Figure Belastung
```{r fig.height=9, fig.width=8}

# labels
sqlabels = read.csv(file=file.path(PATH, 'results', 'sqcodes.txt'), 
                    header = F, sep="\t")
sqlabels$name=c("Violence", "Involvement of children", "Threads", "Severe accidents",
                "Medical emergency", "Dead bodies", "Helplessness", "Contact with family",
                "Suicide", "Time pressure", "Shift work", "Wrong information", "Unclear/no information")

items = names(mydata)[grep('ereignisart_SQ0[0-9][0-9].new$', names(mydata))]
p = list()
for (i in 1:length(items)) {
  
  l = tapply(mydata[[items[i]]], mydata$occupation, summary)
  d = data.frame(matrix(unlist(l), nrow = 5, byrow = T))
  rownames(d) = levels(mydata$occupation)
  colnames(d) = levels(mydata[[items[i]]])
  d=melt(t(d))
  colnames(d) = c("item", "group", "value")
  
  d$Percent = d$value/c(t(matrix(rep(summary(mydata$occupation),6), 5, 3)))
  d$group.short = factor(substr(d$group, 1, 3), levels = c("Pol", "Fir", "Amb", "Eme", "Psy"))
  
  p[[i]] = ggplot(d) + geom_bar(aes(x=group.short, y=Percent, group=item, fill=item), stat = 'identity') + 
    labs(fill="") + theme(axis.text.x = element_text(size=11,angle=90,vjust=0.4), 
                                              legend.text=element_text(size=11)) + xlab("") +
    ggtitle(sqlabels$name[i])
}

plot_grid(plotlist = p, ncol = 3, nrow = 5)
ggsave(path = PATH_FIGURES, filename = "Belastung.png")
```

## Demographics & clinical characteristics
```{r}
# helper functions
mean.sd <- function(x) sprintf("%.1f (%.1f)", mean(x), sd(x))
median.iqr <- function(x, dec=1) sprintf(sprintf("%%.%df (%%.%df)", dec, dec),
                                         median(x, na.rm = T), IQR(x, na.rm = T))

table = data.frame(GroupSize = as.character(summary(mydata$occupation)))
rownames(table)=levels(mydata$occupation)

# Demographics
table$age = tapply(mydata$alter, mydata$occupation, mean.sd)
table$female = round(tapply(mydata$geschlecht=='Frau', mydata$occupation, sum)/summary(mydata$occupation)*100)
table$workExperience = tapply(mydata$dauer, mydata$occupation, mean.sd)
table$relationhip = round(tapply(mydata$zivilstand == "Verheiratet / feste Partnerschaft",
                                 mydata$occupation, sum)/summary(mydata$occupation)*100)
# Clinical variables
table$PTSS = tapply(mydata$PTSS_Skala, mydata$occupation, median.iqr)
table$Suspected = round(tapply(mydata$PSS_Gr == "Verdacht auf PTSD > 12.5", mydata$occupation, sum)/Ng*100)

table$TraumaBefore = round(tapply(mydata$TraumaVorBeruf == ">1 Traumata", mydata$occupation, sum)/Ng*100)
table$TraumaWork = round(tapply(mydata$TraumaArbeit == ">1 Traumata", mydata$occupation, sum)/Ng*100)

table$BSCL_GSI = tapply(mydata$BSCL_GSI, mydata$occupation, median.iqr, 2)
table$BSCL_PST = tapply(mydata$BSCL_PST, mydata$occupation, median.iqr, 1)
table$BSCL_PSDI = tapply(mydata$BSCL_PSDI, mydata$occupation, median.iqr, 2)

table$GHQ = tapply(mydata$GHQ_SUM, mydata$occupation, median.iqr)
table$SWE = tapply(mydata$SWE_Skala, mydata$occupation, median.iqr)

table$Belastung_Beruf =  tapply(mydata$belastung_beruf, mydata$occupation, median.iqr, 2)
table$Belastung_Mensch = tapply(mydata$belastung_menschen, mydata$occupation, median.iqr, 2)

kable(t(table))
```

### Additional global descriptives
```{r}
round(summary(mydata$zivilstand)/N*100)
c(mean(mydata$alter), sd(mydata$alter))
c(mean(mydata$dauer), sd(mydata$dauer))
```


## Figure PTSS
```{r, fig.height=2, fig.width=5}
set.seed(1980)

p1 = ggplot(data=mydata, aes(x=arbeitsort, y=PTSS_Skala, fill=occupation)) +
  geom_boxplot() +
  geom_point(shape=1, color="gray70", size=0.3, position = position_jitter(width = 0.3, height = 0)) +
  theme(axis.text.x=element_blank()) + xlab("occupational group") + ylab("PTSS score")

plot_grid(p1, ncol = 1, nrow = 1, rel_widths = c(1, 1))
ggsave(path = PATH_FIGURES, "fig_PTSS_per_group.png")
```
## Statistican Analysis
### ANOVA
```{r}
fit=aov(PTSS_Skala ~ occupation + dauer + alter + geschlecht, data=mydata)
tab = as.data.frame(summary(fit)[[1]])
tab$`Pr(>F)` = formatPval(tab$`Pr(>F)`)
print(tab)
```

### Verdacht
```{r}
prop.test(tapply(mydata$PSS_Gr == "Verdacht auf PTSD > 12.5", mydata$occupation, sum),
          summary(mydata$occupation))
```

### Belastung Beruf
```{r}
summary(aov(belastung_beruf ~ occupation, data=mydata))
summary(aov(belastung_menschen ~ occupation, data=mydata))
```

## Multiple regression

### Relevel Arbeitsort (Fire Fighters as baseline)
```{r}
mydata$occupation.r = relevel(mydata$occupation, ref = "Fire service")
mydata$geschlecht = relevel(mydata$geschlecht, ref = "Mann")
```


### Model
```{r}
fit1 = lm(PTSS_Skala ~ occupation.r + dauer + geschlecht +
            cope_dysfunct_avoid + cope_dysfunct_subst + cope_dysfunct_dist +
            cope_problfoc_active + cope_problfoc_instr_sup +
            cope_emo_foc +
            TraumaArbeit + TraumaVorBeruf + suizidgedanken +
            SWE_Skala*occupation.r, data=mydata)
summary(fit1)
```

### Table for paper
```{r}
tab = summary(fit1)
idx = order(abs(summary(fit1)$coefficients[,3]), decreasing = TRUE) # sorted by rel. importance, i.e. t-value


tab.nice = tableRegression(fit1, xtable = F)
tab.nice$`t-value` = round(tab$coefficients[,'t value'], digits = 2)
tab.nice.s = tab.nice[idx,c('Coefficient', '95\\%-confidence interval', 't-value', '$p$-value')]
print(tab.nice.s, row.names = F)
```

### Model mit Belastung
```{r}
fit2 = lm(PTSS_Skala ~ occupation.r + dauer + geschlecht +
            cope_dysfunct_avoid + cope_dysfunct_subst + cope_dysfunct_dist +
            cope_problfoc_active + cope_problfoc_instr_sup +
            cope_emo_foc +
            TraumaArbeit + TraumaVorBeruf + suizidgedanken +
            belastung_beruf*occupation.r + belastung_menschen*occupation.r +
            SWE_Skala*occupation.r, data=mydata)

tab = summary(fit2)
idx = order(abs(summary(fit2)$coefficients[,3]), decreasing = TRUE) # sorted by rel. importance, i.e. t-value


tab.nice = tableRegression(fit2, xtable = F)
tab.nice$`t-value` = round(tab$coefficients[,'t value'], digits = 2)
tab.nice.s = tab.nice[idx,c('Coefficient', '95\\%-confidence interval', 't-value', '$p$-value')]
print(tab.nice.s, row.names = F)
```
### Compare models
```{r}
anova(fit1, fit2)
AIC(fit1, fit2)
```

### Collinearity
```{r}
vif(fit1) > 4
```

## Predict suspected PTSD with logistic regression

### Create categories
```{r}
# SWE
mydata$SWE_Skala.cat = rep("Lower", nrow(mydata))
m = median(mydata$SWE_Skala, na.rm = T)
mydata$SWE_Skala.cat[mydata$SWE_Skala > m] = "Upper"
mydata$SWE_Skala.cat = factor(mydata$SWE_Skala.cat, levels = c("Lower", "Upper"))

mydata$dauer.cat = rep('<= 12', nrow(mydata))
m = median(mydata$dauer, na.rm = T)
mydata$dauer.cat[mydata$dauer > m] = "> 12"
mydata$dauer.cat = factor(mydata$dauer.cat, levels = c("<= 12", "> 12"))


# keine wenig etwas stark sehr stark (Ausmass der Belastung)
# 0     1     2     3     4
mydata$belastung_beruf.cat = rep("Low", nrow(mydata))
mydata$belastung_beruf.cat[mydata$belastung_beruf >= 2.5] = "High"
mydata$belastung_beruf.cat = factor(mydata$belastung_beruf.cat, levels = c("Low", "High"))

mydata$belastung_menschen.cat = rep("Low", nrow(mydata))
mydata$belastung_menschen.cat[mydata$belastung_menschen >= 2.5] = "High"
mydata$belastung_menschen.cat = factor(mydata$belastung_menschen.cat, levels = c("Low", "High"))
```

### Model
```{r}
fit3 = glm(PSS_Gr ~ occupation.r + dauer.cat + geschlecht +
             cope_dysfunct_avoid + cope_dysfunct_subst + cope_dysfunct_dist +
             cope_problfoc_active + cope_problfoc_instr_sup + cope_emo_foc +
             anzTraumEreignisse + TraumaVorBeruf +
             belastung_beruf.cat*occupation.r + belastung_menschen.cat*occupation.r +
             SWE_Skala.cat*occupation.r +
             suizidgedanken,
           data = mydata, family = binomial())

tab = summary(fit3)$coefficients

# Create Table
out = as.data.frame(tab)

idx = order(abs(out$`z value`), decreasing = TRUE)

out$`Odds ratio` = exp(out$Estimate)
out$`Odds ratio`= as.character(round(out$Odds, digits = 2))
out$`95% CI` = formatCI(exp(confint(fit3)), text = "english")
out$`Pr(>|z|)` = formatPval(out$`Pr(>|z|)`)
out$`z value` = as.character(signif(out$`z value`, digits = 3))

print(out[idx, c('Odds ratio', '95% CI', 'z value', "Pr(>|z|)")])
```

## SEM
A good model fit is considered as follows:
* CFI ideally larger than 0.9
* RMSEA below 0.10

Model specification

`=~` for latent variables
`~`  causal relationshop
`~~` correlation

Data screening/ SEM assumptions

* Missing values
* normality
* Outliers 

```{r}
# install.packages("lavaan")
library(lavaan)
```

### Model specification

```{r}
model = 'Belastung =~ BSCL_PST + BSCL_GSI + GHQ_Psy + GHQ_Emot
         dysfunctional =~ cope_dysfunct_subst + cope_dysfunct_avoid + cope_dysfunct_dist
         PTSS_Skala ~ dysfunctional + 
                      cope_problfoc_active +
                      SWE_Skala + 
                      TraumaVorBeruf +
                      TraumaArbeit +
                      dauer +
                      geschlecht
         Belastung ~ PTSS_Skala + suizidgedanken'
```



### Fit
```{r}
fit = sem(model=model, data=mydata, group = "occupation")
summary(fit, fit.measures=TRUE, standardized=TRUE)
tab = standardizedSolution(fit)
```
### Table
```{r}
tab$group.name = levels(mydata$occupation)[tab$group]
out = tab[tab$op == "~", c("group.name", "lhs", "rhs", "est.std", "ci.lower", "ci.upper", "pvalue")]

# pretty variable names
out$lhs[out$lhs == "PTSS_Skala"] = "PTSS"
out$lhs[out$lhs == "Belastung"] = "Psychological strain"
out$rhs[out$rhs == "cope_problfoc_active"] = "Active coping"
out$rhs[out$rhs == "SWE_Skala"] = "SE"


# numbers
out$est.std = round(out$est.std, digits = 2)
out$pvalue = formatPval(out$pvalue)
out$CI = formatCI(cbind(out$ci.lower, out$ci.upper), text = "english")

print(out[,  c("group.name", "lhs", "rhs", "est.std", "CI", "pvalue")])

```

### Variance explained
```{r}
tab = as.data.frame(inspect(fit, "rsquare"))
print(round(tab[,c(3,2,1,4,5)], digits = 2))
```





